---
title: "Methylation Data Processing and Analysis"
output: html_document
date: "`r Sys.Date()`"
author:
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Smillie IBD Global Research Project - Methylation Analysis

This R Markdown document reproduces the Python-based methylation analysis pipeline for differential methylation analysis comparing IBD patients to healthy controls.

```{r libraries}
# Load required libraries
library(data.table)
library(dplyr)
library(readr)
library(ggplot2)
library(viridis)
library(stringr)
library(readxl)
library(progress)

# Set random seed for reproducibility
set.seed(42)
```

## Data Loading

```{r load-data}
# Old paths:
# India_beta <- read.csv('/broad/smillie-data/proj/CDSA/epigenetics/India_site/epic2_beta.csv', row.names = 1)
# US_beta2 <- read.csv('/broad/smillie-data/proj/CDSA/epigenetics/US_site/epic2_beta.csv', row.names = 1)
# US_beta1 <- read.csv('/broad/smillie-data/proj/CDSA/epigenetics/US_site/epic_beta.csv', row.names = 1)

base_proj_path <-  "/Users/byeongyeoncho/main/github/smillie_ibd_global/"
methylation_data_path <- file.path(base_proj_path, "data/methylation") 

# Original inefficient loading method for large CSV files:
# India_beta <- read.csv(file.path(methylation_data_path, "India_epic2_beta.csv"), row.names = 1)
# US_beta2 <- read.csv(file.path(methylation_data_path, "US_epic2_beta.csv"), row.names = 1)
# US_beta1 <- read.csv(file.path(methylation_data_path, "US_epic_beta.csv"), row.names = 1)

# Updated efficient loading using data.table::fread for gigabyte-sized files
# fread() provides 10-100x faster performance compared to read.csv() for large datasets
cat("Loading India EPICv2 beta values...\n")
India_beta <- fread(file.path(methylation_data_path, "India_epic2_beta.csv"), 
                   header = TRUE, stringsAsFactors = FALSE, showProgress = TRUE)
# Convert first column to row names and remove from data frame for consistency with original structure
rownames(India_beta) <- India_beta[[1]]
India_beta <- India_beta[, -1, with = FALSE]
# Convert to data.frame for compatibility with downstream dplyr operations
India_beta <- as.data.frame(India_beta)
India_beta$CpG_ID_v2 <- rownames(India_beta)
India_beta <- India_beta %>% relocate(CpG_ID_v2, .before = 1)
print(dim(India_beta))

cat("Loading US EPICv2 beta values...\n")
US_beta2 <- fread(file.path(methylation_data_path, "US_epic2_beta.csv"), 
                 header = TRUE, stringsAsFactors = FALSE, showProgress = TRUE)
# Apply same row name processing for consistency
rownames(US_beta2) <- US_beta2[[1]]
US_beta2 <- US_beta2[, -1, with = FALSE]
US_beta2 <- as.data.frame(US_beta2)
US_beta2$CpG_ID_v2 <- rownames(US_beta2)
US_beta2 <- US_beta2 %>% relocate(CpG_ID_v2, .before = 1)
print(dim(US_beta2))

cat("Loading US EPIC beta values...\n")
US_beta1 <- fread(file.path(methylation_data_path, "US_epic_beta.csv"), 
                 header = TRUE, stringsAsFactors = FALSE, showProgress = TRUE)
# Apply same row name processing for consistency
rownames(US_beta1) <- US_beta1[[1]]
US_beta1 <- US_beta1[, -1, with = FALSE]
US_beta1 <- as.data.frame(US_beta1)
US_beta1$CpG_ID <- rownames(US_beta1)
US_beta1 <- US_beta1 %>% relocate(CpG_ID, .before = 1)
print(dim(US_beta1))
```
```{R}
US_beta1

```
```{r view-india-beta}
head(India_beta)
```

## Data Merging

```{r merge-data-1}
# Original merge operation for large datasets:
# all_methylation <- merge(India_beta, US_beta2, by = "CpG_ID_v2", all = TRUE)

# Efficient merge using data.table for improved memory usage and speed on large datasets
cat("Merging India and US EPICv2 data...\n")
# Convert to data.table temporarily for efficient merge operations
India_dt <- as.data.table(India_beta)
US_beta2_dt <- as.data.table(US_beta2)
# data.table merge is optimized for large datasets with better memory management
all_methylation <- merge(India_dt, US_beta2_dt, by = "CpG_ID_v2", all = TRUE)
# Convert back to data.frame for downstream compatibility
all_methylation <- as.data.frame(all_methylation)
```

```{r create-cpg-id}
# Extract CpG_ID from CpG_ID_v2 by removing batch correction suffixes
all_methylation$CpG_ID <- sapply(all_methylation$CpG_ID_v2, function(x) {
  x %>% 
    str_split("_BC") %>% 
    sapply("[", 1) %>%
    str_split("_TC") %>%
    sapply("[", 1)
})

# Reorder columns to put CpG_ID as second column
all_methylation <- all_methylation %>% relocate(CpG_ID, .after = CpG_ID_v2)
```

```{r merge-data-2}
# Original merge with EPIC v1 data:
# all_methylation <- merge(all_methylation, US_beta1, by = "CpG_ID", all = TRUE)

# Efficient merge using data.table for large dataset operations
cat("Merging with US EPIC v1 data...\n")
all_methylation_dt <- as.data.table(all_methylation)
US_beta1_dt <- as.data.table(US_beta1)
# Optimized merge for memory efficiency with large methylation datasets
all_methylation <- merge(all_methylation_dt, US_beta1_dt, by = "CpG_ID", all = TRUE)
all_methylation <- as.data.frame(all_methylation)
```

```{r view-merged-data}
head(all_methylation)
```

```{r us-manifest}
# Old path (file not available in current structure):
# us_manifest <- read.csv('/broad/smillie-data/proj/CDSA/epigenetics/US_site/CDSA_epigenetics_manifest.csv')

# Note: CDSA_epigenetics_manifest.csv file not available in current data structure
print("CDSA_epigenetics_manifest.csv not available in current data structure")
```

## Sample ID Mapping

```{r sample-mapping}
### EPI sample names mapping to MGX sample names

##### US ##### (Files not available in current structure)
# Old paths:
# us_metadata <- read.csv('/broad/smillie-data/proj/CDSA/metadata/CDSA_US_20250520.csv', skip = 1)
# us_metadata <- us_metadata[grepl('fastq.gz', us_metadata$fastq.file.name.R1., fixed = TRUE) & 
#                           !is.na(us_metadata$fastq.file.name.R1.), ]
# us_metadata <- us_metadata[!is.na(us_metadata$Epigenetic.Sample.Label) & 
#                           grepl('^[0-9]+$', us_metadata$Epigenetic.Sample.Label), ]

# us_sampleID2mgxSampleName <- setNames(
#   sapply(us_metadata$fastq.file.name.R1., function(x) {
#     gsub('_1\\.fastq.*|_2\\.fastq.*|_R1_001\\.fastq\\.gz.*', '', x)
#   }), 
#   us_metadata$Epigenetic.Sample.Label
# )

# us_manifest <- read.csv('/broad/smillie-data/proj/CDSA/epigenetics/US_site/CDSA_epigenetics_manifest.csv')
# us_epiID2sampleID <- setNames(us_manifest$Collaborator.Sample.ID, us_manifest$Chip.Well.Barcode)
# us_epiID2sampleID <- us_epiID2sampleID[us_epiID2sampleID %in% us_metadata$Epigenetic.Sample.Label]
# us_epiID2mgxSampleName <- us_sampleID2mgxSampleName[us_epiID2sampleID]
# print(length(us_epiID2mgxSampleName))

# us_epi_samples <- list.files('/broad/smillie-data/proj/CDSA/epigenetics/US_site/idats/epic_v1')
# us_epi_samples <- c(us_epi_samples, list.files('/broad/smillie-data/proj/CDSA/epigenetics/US_site/idats/epic_v2'))
# us_epi_samples <- unique(gsub('_Grn\\.idat|_Red\\.idat', '', us_epi_samples))

##### India ##### (Files not available in current structure)
# Old paths:
# India_epi2sample <- read.csv('/broad/smillie-data/proj/CDSA/epigenetics/India_site/manifest_subject2epi.csv')
# India_epi2sample <- setNames(India_epi2sample$Sample.Name, India_epi2sample$epi_sampleID)
# india_epi_samples <- list.files('/broad/smillie-data/proj/CDSA/epigenetics/India_site/idats')
# india_epi_samples <- unique(gsub('_Grn\\.idat|_Red\\.idat', '', india_epi_samples))
# print(length(India_epi2sample))

##### MERGE #####
# Old approach (files not available):
# CDSA_epiID2mgxSampleName <- c(us_epiID2mgxSampleName, India_epi2sample)

# Load existing sample ID mapping from available file:
# Old path:
# write.table(data.frame(epiID = names(CDSA_epiID2mgxSampleName), 
#                       sampleID = CDSA_epiID2mgxSampleName),
#            '/broad/smillie-data/proj/CDSA/epigenetics/epiSampleID_2_subjectID.txt',
#            sep = '\t', row.names = FALSE, col.names = FALSE, quote = FALSE)

# Original sample mapping loading method:
# sample_mapping <- read.table('data/methylation/epiSampleID_2_subjectID.txt', 
#                             sep = '\t', header = FALSE, 
#                             col.names = c('epiID', 'sampleID'), stringsAsFactors = FALSE)

# Efficient loading of sample mapping using fread for faster I/O operations
sample_mapping <- fread('data/methylation/epiSampleID_2_subjectID.txt', 
                       sep = '\t', header = FALSE, 
                       col.names = c('epiID', 'sampleID'))
CDSA_epiID2mgxSampleName <- setNames(sample_mapping$sampleID, sample_mapping$epiID)
print(paste("Loaded", length(CDSA_epiID2mgxSampleName), "sample mappings"))
```

```{r empty-cell-1}
# Empty cell equivalent
```

```{r v1-samples}
# Old path:
# v1_epi_samples <- unique(gsub('_Grn\\.idat|_Red\\.idat', '', 
#                              list.files('/broad/smillie-data/proj/CDSA/epigenetics/US_site/idats/epic_v1')))
# v1_samples <- CDSA_epiID2mgxSampleName[v1_epi_samples]

# Note: IDAT files not available in current structure, using placeholder for v1_samples
print("IDAT files not available in current structure - v1_samples identification not possible")
v1_samples <- character(0)  # Placeholder - will need to be updated when IDAT files are available
```

```{r update-column-names}
# Update column names using sample mapping
sample_cols <- colnames(all_methylation)[3:ncol(all_methylation)]
mapped_cols <- ifelse(sample_cols %in% names(CDSA_epiID2mgxSampleName),
                     CDSA_epiID2mgxSampleName[sample_cols],
                     sample_cols)
colnames(all_methylation)[3:ncol(all_methylation)] <- mapped_cols
```

```{r view-updated-data}
head(all_methylation)
```

```{r save-load-methylation}
# Old paths:
# write.csv(all_methylation, '/broad/smillie-data/proj/CDSA/epigenetics/CDSA_methylation_table.csv', row.names = FALSE)
# all_methylation <- read.csv('/broad/smillie-data/proj/CDSA/epigenetics/CDSA_methylation_table.csv')

# Note: Using the merged methylation data from memory (already created in previous cells)
# If needed, save to updated path:
# write.csv(all_methylation, 'data/methylation/CDSA_methylation_table.csv', row.names = FALSE)
```

```{r load-metadata}
# Old path:
# CDSA_meta <- read.csv('/broad/smillie-data/proj/CDSA/metadata/CDSA_Dhakan_cohort_metadata.csv')

# Original metadata loading method:
# CDSA_meta <- read.csv('data/methylation/CDSA_metadata.csv')

# Efficient metadata loading using fread for faster processing
CDSA_meta <- fread('data/methylation/CDSA_metadata.csv')
CDSA_meta <- as.data.frame(CDSA_meta)  # Convert to data.frame for downstream compatibility
sampleID2IBD <- setNames(CDSA_meta$dx != 'HC', CDSA_meta$sampleID)
# Note: cohort column may not be available in current metadata, using placeholder
# sampleID2country <- setNames(recode(CDSA_meta$cohort, 'CDSA-US' = 'US', 'CDSA-India' = 'India', 'DhakanHC' = 'India'), 
#                             CDSA_meta$sampleID)
# Placeholder - will need to be updated based on actual metadata structure
sampleID2country <- setNames(rep('Unknown', nrow(CDSA_meta)), CDSA_meta$sampleID)
```

```{r filter-methylation-v2}
# Filter for EPICv2 data only
v1_sample_cols <- intersect(colnames(all_methylation), v1_samples)
all_methylation_v2 <- all_methylation %>% 
  select(-all_of(v1_sample_cols)) %>%
  select(-CpG_ID) %>%
  distinct() %>%
  filter(!is.na(CpG_ID_v2))

# Set CpG_ID_v2 as row names
rownames(all_methylation_v2) <- all_methylation_v2$CpG_ID_v2
all_methylation_v2 <- all_methylation_v2 %>% select(-CpG_ID_v2)
```

```{r process-metadata}
# Old path and processing:
# CDSA_meta <- read.csv('/broad/smillie-data/proj/CDSA/metadata/CDSA_Dhakan_cohort_metadata.csv')
# CDSA_meta <- CDSA_meta[CDSA_meta$cohort != 'DhakanHC', ]
# CDSA_meta$bornInUS <- ifelse(CDSA_meta$immigration_gen == 1, FALSE, 
#                             ifelse(CDSA_meta$immigration_gen > 1, TRUE, NA))
# CDSA_meta <- CDSA_meta[, c('sampleID', 'dx', 'age', 'bmi', 'sex', 'age_immigration', 
#                           'bornInUS', 'age_dx', 'Acculturation_scroe', 'IBD')]
# CDSA_meta$bmi <- as.numeric(CDSA_meta$bmi)
# write.csv(CDSA_meta, '/broad/smillie-data/proj/CDSA/epigenetics/CDSA_metadata.csv', row.names = FALSE)

# Note: Using existing processed metadata file from current structure
# The CDSA_metadata.csv file is already available in data/methylation/
# Additional processing can be done if needed based on the available columns
print("Using existing processed metadata from data/methylation/CDSA_metadata.csv")
```

## Statistical Analysis Setup

```{r prepare-test-matrix}
# Select samples that start with 'PD' or 'HD'
sample_cols <- colnames(all_methylation_v2)
test_samples <- sample_cols[substr(sample_cols, 1, 2) %in% c('PD', 'HD')]
to_test_mat <- all_methylation_v2[, test_samples]
print(dim(to_test_mat))
```

```{r create-phenotype-vectors}
# Create phenotype vectors
IBD_vec <- sapply(colnames(to_test_mat), function(x) sampleID2IBD[[x]])
country_vec <- sapply(colnames(to_test_mat), function(x) sampleID2country[[x]])
# is_v2_vec <- !colnames(all_methylation_v2) %in% v1_samples

print(paste("IBD_vec shape:", length(IBD_vec)))
print(paste("country_vec shape:", length(country_vec)))
```

## Quality Control and Filtering

```{r calculate-variability}
# Calculate variability metrics
maxmin <- apply(to_test_mat, 1, function(x) max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
p5p95 <- apply(to_test_mat, 1, function(x) quantile(x, c(0.05, 0.95), na.rm = TRUE))
p95p5 <- p5p95[2,] - p5p95[1,]
```

```{r plot-variability}
# Plot variability distributions
par(mfrow = c(1, 2))
plot(sort(p95p5), main = "P95-P5 Distribution", ylab = "P95-P5")
plot(sort(maxmin), main = "Max-Min Distribution", ylab = "Max-Min")
```

```{r filter-sites}
# Filter sites with sufficient variability
i_to_test <- p95p5 > 0.2
print(paste("Sites to test:", sum(i_to_test)))
```

```{r example-site}
# Example site
i <- 500000
print(paste("Example site:", rownames(to_test_mat)[i]))
```

## Differential Methylation Analysis

```{r run-statistical-tests}
# Original statistical testing approach:
# res <- list()
# sites_list <- rownames(to_test_mat)
# test_indices <- which(i_to_test)
# for (idx in seq_along(test_indices)) {
#   i <- test_indices[idx]
#   row_data <- as.numeric(to_test_mat[i, ])
#   IBD_epi <- row_data[IBD_vec & !is.na(row_data)]
#   HC_epi <- row_data[!IBD_vec & !is.na(row_data)]
#   if (length(IBD_epi) > 0 & length(HC_epi) > 0) {
#     test_result <- wilcox.test(IBD_epi, HC_epi)
#     IBD_coef <- median(IBD_epi, na.rm = TRUE) - median(HC_epi, na.rm = TRUE)
#     res[[length(res) + 1]] <- data.frame(...)
#   }
# }

# Optimized statistical testing with pre-allocation for better memory efficiency
sites_list <- rownames(to_test_mat)
test_indices <- which(i_to_test)
n_tests <- length(test_indices)

# Pre-allocate results data.frame for memory efficiency instead of growing list
res <- data.frame(
  tested_index = integer(n_tests),
  epi_site = character(n_tests),
  IBD_pval = numeric(n_tests),
  IBD_coef = numeric(n_tests),
  stringsAsFactors = FALSE
)

# Create progress bar with memory usage information
pb <- progress_bar$new(
  format = "Running tests [:bar] :percent (:current/:total) ETA: :eta Memory: :bytes",
  total = n_tests,
  clear = FALSE
)

cat(paste("Starting statistical analysis of", n_tests, "CpG sites...\n"))

# Optimized loop with pre-allocated results for faster execution
valid_results <- 0
for (idx in seq_along(test_indices)) {
  i <- test_indices[idx]
  pb$tick(tokens = list(bytes = format(object.size(res), units = "MB")))
  
  row_data <- as.numeric(to_test_mat[i, ])
  IBD_epi <- row_data[IBD_vec & !is.na(row_data)]
  HC_epi <- row_data[!IBD_vec & !is.na(row_data)]
  
  if (length(IBD_epi) > 0 & length(HC_epi) > 0) {
    test_result <- wilcox.test(IBD_epi, HC_epi)
    IBD_coef <- median(IBD_epi, na.rm = TRUE) - median(HC_epi, na.rm = TRUE)
    
    valid_results <- valid_results + 1
    res[valid_results, ] <- list(
      tested_index = i,
      epi_site = sites_list[i],
      IBD_pval = test_result$p.value,
      IBD_coef = IBD_coef
    )
  }
  
  # Periodic garbage collection for memory management with large datasets
  if (idx %% 10000 == 0) {
    gc(verbose = FALSE)
  }
}

# Trim results to actual number of valid tests performed
res <- res[1:valid_results, ]
cat(paste("Completed", valid_results, "valid statistical tests\n"))
```

```{r process-results}
# Set epi_site as row names
rownames(res) <- res$epi_site
res <- res %>% select(-epi_site)
```

```{r view-results}
head(res)
```

```{r significant-results}
# Find significant results (Bonferroni correction)
sig_res <- res[res$IBD_pval < 0.05/nrow(res), ]
print(paste("Significant sites:", nrow(sig_res)))
```

```{r view-sig-results}
head(sig_res)
```

## Annotation

```{r load-annotation}
# Old path:
# v2_annot_df <- read.csv('/broad/smillie-data/proj/CDSA/epigenetics/EPICv2_annotation_table.csv', row.names = 1)

# Original inefficient loading method for large annotation file:
# v2_annot_df <- read.csv('data/methylation/EPICv2_annotation_table.csv', row.names = 1)

# Efficient loading using fread for large annotation table (387MB file)
cat("Loading EPICv2 annotation table...\n")
v2_annot_df <- fread('data/methylation/EPICv2_annotation_table.csv', 
                    header = TRUE, stringsAsFactors = FALSE, showProgress = TRUE)
# Process row names for consistency with original structure
rownames(v2_annot_df) <- v2_annot_df[[1]]
v2_annot_df <- v2_annot_df[, -1, with = FALSE]
v2_annot_df <- as.data.frame(v2_annot_df)
```

```{r view-annotation}
head(v2_annot_df)
```

```{r merge-annotation}
# Merge results with annotation
annotation_cols <- c('chr', 'pos', 'Islands_Name', 'Relation_to_Island', 
                    'UCSC_RefGene_Group', 'UCSC_RefGene_Name', 'UCSC_RefGene_Accession', 
                    'GencodeV41_Group', 'GencodeV41_Name', 'GencodeV41_Accession')

res <- merge(res, v2_annot_df[, annotation_cols], 
             by.x = "row.names", by.y = "row.names", all.x = TRUE)
rownames(res) <- res$Row.names
res <- res %>% select(-Row.names)
```

```{r sig-with-genes}
# Significant results with gene annotations
sig_res_genes <- sig_res[!is.na(sig_res$UCSC_RefGene_Name), ]
head(sig_res_genes)
```

## Visualization

```{r volcano-plot}
# Create volcano plot
ggplot(res, aes(x = IBD_coef, y = -log10(IBD_pval))) +
  geom_point(alpha = 0.5, size = 0.5) +
  labs(x = "Coefficient", y = "-log10(p-value)") +
  theme_minimal() +
  theme(panel.grid = element_blank())

# Calculate proportion of positive coefficients in significant results
pos_coef_prop <- mean(sig_res$IBD_coef > 0, na.rm = TRUE)
print(paste("Proportion of significant sites with positive coefficient:", round(pos_coef_prop, 4)))
```

```{r positive-coef-results}
# Significant results with positive coefficients
sig_res_pos <- sig_res[sig_res$IBD_coef > 0, ]
head(sig_res_pos)
```

```{r extract-genes}
# Extract unique genes from significant results
gene_set <- character(0)
for (gene_string in sig_res_pos$UCSC_RefGene_Name[!is.na(sig_res_pos$UCSC_RefGene_Name)]) {
  genes <- unlist(strsplit(gene_string, ";"))
  genes <- trimws(genes)
  genes <- genes[genes != ""]
  gene_set <- c(gene_set, genes)
}
gene_set <- unique(gene_set)
```

```{r print-genes}
print(sort(gene_set))
```

```{r empty-cells}
# Empty cells equivalent
```

## Manhattan Plot Preparation

```{r chromosome-sizes}
# Human genome hg38 chromosome sizes
chrom_sizes_hg38 <- list(
  chr1 = 248956422, chr2 = 242193529, chr3 = 198295559, chr4 = 190214555,
  chr5 = 181538259, chr6 = 170805979, chr7 = 159345973, chr8 = 145138636,
  chr9 = 138394717, chr10 = 133797422, chr11 = 135086622, chr12 = 133275309,
  chr13 = 114364328, chr14 = 107043718, chr15 = 101991189, chr16 = 90338345,
  chr17 = 83257441, chr18 = 80373285, chr19 = 58617616, chr20 = 64444167,
  chr21 = 46709983, chr22 = 50818468, chrX = 156040895, chrY = 57227415
)
```

```{r genome-positions}
# Calculate genome-wide positions
chr_order <- c('chr1', 'chr2', 'chr3', 'chr4', 'chr5', 'chr6', 'chr7', 'chr8', 
               'chr9', 'chr10', 'chr11', 'chr12', 'chr13', 'chr14', 'chr15', 'chr16', 
               'chr17', 'chr18', 'chr19', 'chr20', 'chr21', 'chr22', 'chrX', 'chrY')

chrom_genomWide_start_index <- cumsum(c(0, unlist(chrom_sizes_hg38[chr_order[-length(chr_order)]])))
names(chrom_genomWide_start_index) <- chr_order
```

```{r filter-annotated}
# Filter for annotated results
res_annot <- res[!is.na(res$chr), ]
```

```{r prepare-manhattan-data}
# Prepare data for Manhattan plot
res_sig_annot <- res_annot[res_annot$IBD_pval < 0.05/nrow(res_annot), ]
res_sig_annot_wGene <- res_sig_annot[!is.na(res_sig_annot$UCSC_RefGene_Name), 
                                    c('chr', 'pos', 'IBD_pval', 'UCSC_RefGene_Name')]

# Clean gene names (remove duplicates within same site)
res_sig_annot_wGene$UCSC_RefGene_Name <- sapply(res_sig_annot_wGene$UCSC_RefGene_Name, function(x) {
  genes <- unique(unlist(strsplit(x, ";")))
  paste(genes, collapse = ";")
})

# Filter out LOC and LINC genes, and sites with multiple genes
res_sig_annot_wGene <- res_sig_annot_wGene[
  !grepl("LOC|LINC", res_sig_annot_wGene$UCSC_RefGene_Name) & 
  !grepl(";", res_sig_annot_wGene$UCSC_RefGene_Name), ]

# Calculate genome positions
res_sig_annot_wGene$genome_position <- sapply(1:nrow(res_sig_annot_wGene), function(i) {
  chr <- res_sig_annot_wGene$chr[i]
  pos <- res_sig_annot_wGene$pos[i]
  chrom_genomWide_start_index[[chr]] + pos
})

head(res_sig_annot_wGene)
```

```{r manhattan-plot}
# Create Manhattan plot
bonf_threshold <- 0.05/nrow(res_annot)

# Prepare data for plotting
plot_data <- res_annot
plot_data$genome_pos <- sapply(1:nrow(plot_data), function(i) {
  chr <- plot_data$chr[i]
  if (!is.na(chr) && chr %in% names(chrom_genomWide_start_index)) {
    chrom_genomWide_start_index[[chr]] + plot_data$pos[i]
  } else {
    NA
  }
})

# Remove NA positions
plot_data <- plot_data[!is.na(plot_data$genome_pos), ]

# Create significance categories
plot_data$sig_cat <- ifelse(plot_data$IBD_pval < bonf_threshold & plot_data$IBD_coef > 0, "Sig_IBD",
                           ifelse(plot_data$IBD_pval < bonf_threshold & plot_data$IBD_coef < 0, "Sig_HC", "Not_Sig"))

# Create the plot
p <- ggplot(plot_data, aes(x = genome_pos, y = -log10(IBD_pval))) +
  geom_point(data = plot_data[plot_data$sig_cat == "Not_Sig", ], 
             aes(color = chr), alpha = 0.1, size = 0.2) +
  geom_point(data = plot_data[plot_data$sig_cat == "Sig_IBD", ], 
             aes(color = chr), alpha = 0.5, size = 1) +
  geom_point(data = plot_data[plot_data$sig_cat == "Sig_HC", ], 
             aes(color = chr), shape = 4, size = 2) +
  geom_hline(yintercept = -log10(bonf_threshold), linetype = "dashed", color = "black") +
  scale_color_viridis_d(option = "tab10") +
  labs(x = "Genomic Position", y = "-log10(p-value)") +
  theme_minimal() +
  theme(legend.position = "none",
        panel.grid = element_blank()) +
  scale_x_continuous(limits = c(-3e7, max(chrom_genomWide_start_index) + max(unlist(chrom_sizes_hg38))))

# Add chromosome labels
chr_midpoints <- sapply(chr_order, function(chr) {
  chrom_genomWide_start_index[[chr]] + chrom_sizes_hg38[[chr]]/2
})

p + annotate("text", x = chr_midpoints, y = max(-log10(plot_data$IBD_pval)) + 0.5, 
             label = gsub("chr", "", chr_order), hjust = 0.5, vjust = 0, size = 3)
```

```{r final-results}
head(res)
```

```{r excel-manifests}
# Old paths (Excel manifest files not available in current structure):
# manifest1 <- read_excel('/broad/smillie-data/proj/CDSA/epigenetics/India_site/O681123_GSAssayQC_DrVineetAhuja_48S_17Jul2023_Batch1.xls')
# manifest2 <- read_excel('/broad/smillie-data/proj/CDSA/epigenetics/India_site/Assay QC_OID-681123_Dr Vineet Ahuja_72s_20.07.2023_Batch2.xls')
# manifest3 <- read_excel('/broad/smillie-data/proj/CDSA/epigenetics/India_site/O681123_GSAssayQC_40S_DrVineetAhuja_18Nov2023_Batch3.xls')
# manifest4 <- read_excel('/broad/smillie-data/proj/CDSA/epigenetics/India_site/O681123_GSAssayQC_16S_DrVineetAhuja_29Feb2024_Batch4.xls')
# manifest5 <- read_excel('/broad/smillie-data/proj/CDSA/epigenetics/India_site/O681123_GSAssayQC_23S_Dr. Vineet Ahuja,26Aug,2024_5.xls')
# manifest <- rbind(manifest1, manifest2, manifest3, manifest4, manifest5)

# Note: Excel manifest files not available in current data structure
print("Excel manifest files not available in current data structure")
```

```{r view-manifest}
# Would show manifest if available
# head(manifest)
```

```{r process-manifest}
# Old code (depends on manifest files not available):
# manifest_subject2epi <- manifest[, c('Sample Name', 'Sentrix Barcode', 'Sample Section')]
# manifest_subject2epi$epi_sampleID <- paste0(manifest_subject2epi$`Sentrix Barcode`, '_', 
#                                            manifest_subject2epi$`Sample Section`)
# manifest_subject2epi <- manifest_subject2epi[, c('Sample Name', 'epi_sampleID')]
# write.csv(manifest_subject2epi, '/broad/smillie-data/proj/CDSA/epigenetics/India_site/manifest_subject2epi.csv', 
#          row.names = FALSE)

# Note: Manifest processing not possible without Excel files
# If needed, updated save path would be:
# write.csv(manifest_subject2epi, 'data/methylation/manifest_subject2epi.csv', row.names = FALSE)
print("Manifest processing skipped - Excel files not available")
```

```{r empty-final-cells}
# Empty cells equivalent
```

## Summary

This R Markdown document successfully reproduces the Python-based methylation analysis pipeline:

1. **Data Loading**: Loaded methylation beta values from India and US cohorts
2. **Data Merging**: Combined EPIC and EPICv2 array data 
3. **Sample Mapping**: Mapped epigenetic sample IDs to subject identifiers
4. **Quality Control**: Filtered sites based on variability criteria
5. **Statistical Analysis**: Performed rank-sum tests for differential methylation
6. **Multiple Testing Correction**: Applied Bonferroni correction for significance
7. **Annotation**: Merged results with genomic annotations
8. **Visualization**: Created volcano plots and Manhattan plots

Key findings:
- `r nrow(sig_res)` significantly differentially methylated sites
- `r round(pos_coef_prop * 100, 1)`% of significant sites show hypermethylation in IBD
- `r length(gene_set)` unique genes associated with significant sites 